---
title: "Anti-Patterns"
engine: knitr
filters:
  - interactive-duckdb
databases:
  - name: cookbook
    path: "https://raw.githubusercontent.com/Alaburda/data-analyst-sql-cookbook/master/databases/carwash.duckdb"
    format: duckdb
---

```{r setup}
#| echo: false
#| message: false
#| warning: false

library(duckdb)
library(DBI)

con <- dbConnect(duckdb(), "../databases/carwash.duckdb", read_only = TRUE)

# Render a query alongside its DuckDB physical plan as two clean code blocks.
# DuckDB's EXPLAIN returns an ASCII tree; printing it through a normal SQL chunk
# turns it into an HTML <table> that mangles the alignment. Emitting it as raw
# markdown (results: asis) keeps it in a monospaced <pre> block instead.
# `pre` holds statements to run *and* display first, e.g. a SET VARIABLE.
show_plan <- function(query, pre = NULL) {
  display <- query
  if (!is.null(pre)) {
    for (s in pre) dbExecute(con, s)
    display <- paste0(paste0(pre, ";\n\n", collapse = ""), query)
  }
  plan <- dbGetQuery(con, paste("EXPLAIN", query))[["explain_value"]]
  cat(sprintf("```sql\n%s\n```\n\n::: {.sql-plan}\n```\n%s\n```\n:::\n",
              display, plan))
}

```

```{=html}
<style>
.sql-plan { overflow-x: auto; }
.sql-plan pre { white-space: pre; font-size: 0.78rem; line-height: 1.2; }
</style>
```

SQL is a sneaky language because you can keep building your select statement to your heart's desires until one change suddenly makes the query slow. There are some anti-patterns you can avoid - in fact, they're easy to spot with enough practice.

## Doing dbt style statements for the latest values  

So dbt has been recommending doing incremental refreshes using this sort of syntax:

```{r}
#| echo: false
#| results: asis
show_plan(
"SELECT *
FROM subscriptions
WHERE date_from >= (SELECT max(date_from) FROM subscriptions)"
)
```

Notice the plan: the scalar subquery forces a `NESTED_LOOP_JOIN` against a second scan of `subscriptions`, plus a `CASE WHEN … error('More than one row…')` guard that DuckDB injects to enforce scalar-subquery semantics. This happens because the database cannot use an existing query plan and can't do "parameter sniffing". One way to avoid this is to set the max value from the subquery and then pass it forward to the main query:

```{r}
#| echo: false
#| results: asis
show_plan(
"SELECT *
FROM subscriptions
WHERE date_from >= getvariable('max_date')",
  pre = "SET VARIABLE max_date = (SELECT max(date_from) FROM subscriptions)"
)
```

Now the plan collapses to a single `SEQ_SCAN` with a constant filter — the subquery is evaluated once, up front, and the main query never sees it.


## Using OR in joins

Just because you can use an OR statement in a SQL query, doesn't mean you have to - in practice, the query optimizer may not correctly [interpret what you're doing](https://stackoverflow.com/questions/5901791/is-having-an-or-in-an-inner-join-condition-a-bad-idea). See the query plan for this:

```{r}
#| echo: false
#| results: asis
show_plan(
"SELECT users.* 
FROM users 
INNER JOIN calendar cal
  ON signup_date = cal.date
  OR confirmation_date = cal.date
where year(cal.date) = 2023")
```

```{r}
#| echo: false
#| results: asis
show_plan(
"SELECT users.* 
FROM users 
INNER JOIN calendar cal
  ON signup_date = cal.date
where year(cal.date) = 2023

UNION 

SELECT users.* 
FROM users 
INNER JOIN calendar cal
  ON confirmation_date = cal.date
where year(cal.date) = 2023
")



```

`BLOCKWISE_NL_JOIN` in the query plan is a blockwise nested loop join - it's a very inefficient algorithm whereas the UNION is much faster.

## left join with left side predicate in join 

I haven't tried this in other databases, but apparently DuckDB doesn't support efficient joins when you want to join a table only to some of the rows but you want to keep the whole table intact. Luckily, in my experience, this is easy to overcome using a `case when` statement and overall this pattern is VERY rare! 

```{r}

#| echo: false
#| results: asis
show_plan(
"SELECT users.* 
FROM users 
LEFT JOIN calendar cal
  ON confirmation_date = cal.date
  AND year(confirmation_date) = 2023
")

```


```{sql, connection = con}
#| eval: false

EXPLAIN


```

```{sql, connection = con}
#| eval: false

EXPLAIN
WITH tmp AS (
  SELECT *,
    CASE WHEN plan_autorenew = 1 THEN 1 WHEN plan_legacy = 1 THEN 1 END AS plan_vip
  FROM subscriptions
)
SELECT *
FROM tmp
INNER JOIN vip ON vip.id = plan_vip

```

```{sql, connection = con}
#| eval: false

EXPLAIN
SELECT * FROM subscriptions INNER JOIN vip ON plan_autorenew = vip.id AND plan_legacy IS NULL
UNION ALL
SELECT * FROM subscriptions INNER JOIN vip ON plan_legacy = vip.id AND plan_autorenew IS NULL

```

## Using UNION where UNION ALL is sufficient

UNION deduplicates data after the join, which is a waste of resources if you know that the data is already unique. Use UNION ALL instead.

```{r}
#| echo: false
#| results: asis
show_plan(
"SELECT users.* 
FROM users 
INNER JOIN calendar cal
  ON signup_date = cal.date
where year(cal.date) = 2023

UNION 

SELECT users.* 
FROM users 
INNER JOIN calendar cal
  ON confirmation_date = cal.date
where year(cal.date) = 2023
")



```

```{r}
#| echo: false
#| results: asis
show_plan(
"SELECT users.* 
FROM users 
INNER JOIN calendar cal
  ON signup_date = cal.date
where year(cal.date) = 2023

UNION ALL

SELECT users.* 
FROM users 
INNER JOIN calendar cal
  ON confirmation_date = cal.date
where year(cal.date) = 2023
")



```

## GROUP BY vs. DISTINCT vs. EXISTS

EXISTS is the fastest but also most verbose way of returning unique values in a table. See the following:




